---
title: "B2_analysis_with_BAMA"
author: "Fan Bu"
date: "3/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = '~/Documents/Research_and_References/HIV_rebound_summer2020/')
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(survival)
library(survminer)
library(glmnet)
library(stringr)
```


## Multivariate analysis with BAMA data
Load data.
```{r}
#dat_log = readRDS('reboundB2_logTrans_ADCC_ADCP.rds')
# update version: include more timepoints of measurements
# for GP41, GP120, ADCC & ADCP
dat_log = readRDS('reboundB2_logTrans_032021.rds')
bama.dat = readRDS('BAMA_data_clean.rds')
```



Select the covariates required for the re-analysis from the previous dataset.
```{r}
#sel = c(1:11, 21:23, 43:44, 46:48)
sel = c(1:11, 21:23, 43:44, 46:52) #include more covars, March 2021
dat_log_sel = dat_log[,sel]
```

Then combine the dataset with select BAMA variables
```{r}
bama_sel = bama.dat %>% arrange(animal_id) %>%
  select(SHIV.CH505TF.V2_week8, SHIV.CH505TF.V2_ATI,
         SHIV.CH505TF.V1V2_week8, SHIV.CH505TF.V1V2_ATI,
         SHIV.CH505TF.V3_week8, SHIV.CH505TF.V3_ATI,
         SIVmac239.p27_week8, SIVmac239.p27_ATI,
         SIVmac251.Nef_week8, SIVmac251.Nef_ATI)
dat_log_sel = cbind(dat_log_sel, bama_sel)
```

Make the "Sex" and "A01" columns numeric so that we can run LASSO.
```{r}
# March 18, 2021 update: Sex baseline to Female (F=0, M=1)
dat_log_sel = dat_log_sel %>% 
  mutate(Sex = if_else(Sex=='F', 0, 1),
         A01 = if_else(A01=='Neg', 0, 1))
```

Run Cox PH with Lasso.
(Also tried a version without adjusting for Sex and A01...)
```{r}
#X = dat_log_sel[,6:29] %>% as.matrix()

X = dat_log_sel[,6:33] %>% as.matrix()

# take out Sex and A01...
# X = dat_log_sel[,6:29][,-c(10,11)] %>% as.matrix()

phmod_lasso = glmnet(X, 
                     Surv(dat_log_sel$rebound_time_days_post_ati, 
                          dat_log_sel$observed),
                     family = "cox")
```

Get a "predictor inclusion" ranking table.
```{r, message=FALSE, warning=FALSE}
set.seed(42)

# get inclusion order
pred_in = NULL
coef_sign = NULL

# stop when num. of active predictors is > 7
for(l in phmod_lasso$lambda){
  coefficients = coef(phmod_lasso, s = l)
  active_index = which(coefficients != 0)
  active_coefficients = coefficients[active_index]
  active_predictors = attr(coefficients,"Dimnames")[[1]][active_index]
  
  #cat(active_predictors, "\n")
  if(length(active_index) > 7){ break }
  
  if(any(!active_predictors %in% pred_in)){
    #cat('Added new predictors! Now active:', active_predictors, '\n')
    
    new_index = which(!active_predictors %in% pred_in)
    pred_in = c(pred_in, active_predictors[new_index])
    new_coef_sign = ifelse(active_coefficients[new_index] > 0, 
                           "+", "-")
    coef_sign = c(coef_sign, new_coef_sign)
  }
}

# focus on those that are eventually selected...
#coef_sign = coef_sign[pred_in %in% active_predictors]
#pred_in = pred_in[pred_in %in% active_predictors]

# rebound effect
rebound_effect = sapply(coef_sign, 
                        function(x) ifelse(x=="+","accelerate","delay")) %>%
  as.vector()

# multivariate concordance (C-statistic)
Response = "Surv(rebound_time_days_post_ati, observed)"

Cum_C_stats = NULL
for(i in 1:length(pred_in)){
  covars = pred_in[1:i]
  f = as.formula(paste(Response,paste(covars,collapse = "+"),sep = " ~ "))
  mod = coxph(f, data=dat_log_sel)
  C_covars = mod$concordance['concordance'] %>% as.numeric()
  Cum_C_stats = c(Cum_C_stats, C_covars)
}

# leave-one-out CV error
cv_phmod_lasso3 = cv.glmnet(X, 
                            Surv(dat_log_sel$rebound_time_days_post_ati, 
                                 dat_log_sel$observed),
                            family = "cox", nfolds = 10, keep = T)
pred_in2 = NULL
LOO_deviance = NULL
for(l in phmod_lasso$lambda){
  coefficients = coef(phmod_lasso, s = l)
  active_index = which(coefficients != 0)
  #active_coefficients = coefficients[active_index]
  active_predictors = attr(coefficients,"Dimnames")[[1]][active_index]
  
  #cat(active_predictors, "\n")
  
  if(length(active_index) > 7){ break }
  
  if(any(!active_predictors %in% pred_in2)){
    new_index = which(!active_predictors %in% pred_in2)
    pred_in2 = c(pred_in2, active_predictors[new_index])
    
    l_index = which(cv_phmod_lasso3$lambda == l)
    LOO_deviance = c(LOO_deviance, cv_phmod_lasso3$cvm[l_index])
  }
}

#LOO_deviance = LOO_deviance[pred_in2 %in% active_predictors]
#pred_in2 = pred_in2[pred_in2 %in% active_predictors]
```

Put together a summary table.
```{r}
predictor_inclusion = data.frame(Inclusion_Rank = c(1:length(pred_in)),
                                 Predictor = pred_in,
                                 #Coefficient = coef_sign,
                                 Rebound_Effect = rebound_effect,
                                 Cum_concordance = Cum_C_stats,
                                 LOOCV_deviance = LOO_deviance)
predictor_inclusion
```

Still, the most important predictor (first to include) is `Pos AUC at ATI`.

Save the inclusion ranking table to a txt file for Word usage
```{r}
write.table(format(predictor_inclusion, digits=3), 
            'predictor_inclusion_table.txt',
            quote=FALSE, row.names = FALSE,
            sep = ',')
```


Visualize the effect by KM curves.
(Chunk below is deleted as `Sex` is no longer the 2nd predictor.)
```{r}
phmod_auc0_Sex = coxph(Surv(rebound_time_days_post_ati, observed) ~ 
                          pos_auc_0_weeks_post_ATI + Sex, 
                        data = dat_log_sel)
auc0_values = c(0.15, 0.25, 0.3, 0.4) # 0.25 is approximately the mean

# Sex=female first
auc0_fit = survfit(phmod_auc0_Sex, 
                   newdata = data.frame(pos_auc_0_weeks_post_ATI = auc0_values,
                                        Sex = 0)) 
ggsurvplot(auc0_fit, conf.int = FALSE, 
           data = dat_log,
           legend = "right",
           legend.title = "Pos AUC \n0 weeks\npost ATI",
           legend.labs=c("0.15", 
                         "0.25(mean)",
                         "0.30",
                         "0.40"),
           caption = "Fix Sex as Female",
           ggtheme = theme_bw(base_size = 14))

# Then Sex=male
auc0_fit = survfit(phmod_auc0_Sex, 
                   newdata = data.frame(pos_auc_0_weeks_post_ATI = auc0_values,
                                        Sex = 1)) 
ggsurvplot(auc0_fit, conf.int = FALSE, 
           data = dat_log,
           legend = "none",
           caption = "Fix Sex as Male",
           ggtheme = theme_bw(base_size = 14))
```

Check out the model using `Pos_AUC_0` and `ADCP_week8`.

Effect of `Pos_AUC_0` is significant but effect of `ADCP_week8` is **not**.

```{r}
phmod_auc0_adcp = coxph(Surv(rebound_time_days_post_ati, observed) ~
                          pos_auc_0_weeks_post_ATI + ADCP_week8, 
                        data = dat_log_sel)
summary(phmod_auc0_adcp)
```

Plot effect of `pos_auc_0` when fixing `ADCP_week8` at mean value.

```{r}
mean(dat_log_sel$ADCP_week8) # 14.5878

auc0_values = c(0.15, 0.25, 0.3, 0.4) # 0.25 is approximately the mean

auc0_fit = survfit(phmod_auc0_adcp, 
                   newdata = data.frame(
                     pos_auc_0_weeks_post_ATI = auc0_values,
                     ADCP_week8 = mean(dat_log_sel$ADCP_week8)))
ggsurvplot(auc0_fit, conf.int = FALSE, 
           data = dat_log_sel,
           legend = "right",
           legend.title = "Pos AUC \n0 weeks\npost ATI",
           legend.labs=c("0.15", 
                         "0.25(mean)",
                         "0.30",
                         "0.40"),
           caption = "Fix ADCP_week8 at 14.6 (mean)",
           ggtheme = theme_bw(base_size = 14))
```

Then plot effect of `ADCP_week8` while fixing `Pos_AUC_0` at mean value.
```{r}
#sort(dat_log_sel$ADCP_week8)

adcp_values = c(11, 14.6, 17, 19) # 14.6 is approx. the mean

adcp_fit = survfit(phmod_auc0_adcp, 
                   newdata = data.frame(
                     pos_auc_0_weeks_post_ATI =
                       mean(dat_log_sel$pos_auc_0_weeks_post_ATI),
                     ADCP_week8 = adcp_values))
ggsurvplot(adcp_fit, conf.int = FALSE, 
           data = dat_log_sel,
           legend = "right",
           legend.title = "ADCP(%) \n0 Week 8",
           legend.labs=c("11.0", 
                         "14.6(mean)",
                         "17.0",
                         "19.0"),
           caption = "Fix Pos_AUC_0 at 0.25 (mean)",
           ggtheme = theme_bw(base_size = 14))
```




## Run Multivariate Cox PH model with Lasso for variable selection using ALL biomarker predictors used in previous analysis

Include all variables used in previous analysis (the antibody response covariates).
```{r}
dat_log_prev = dat_log %>%
  mutate(Sex = if_else(Sex=='F', 0, 1),
         A01 = if_else(A01=='Neg', 0, 1)) %>%
  select(rebound_time_days_post_ati, observed, 
         Challenge_times, Dosage, 
         Sex, A01, 
         log_peak_gp41, log_peak_gp120,
         log_gp41_treat, log_gp120_treat,
         log_point_ic50_0_weekspost_ATI:log_point_ic50_8_weekspost_ATI,
         pos_auc_0_weeks_post_ATI:pos_auc_8_weeks_post_ATI,
         ADCC_week56:ADCP_week8, 
         log_gp41_ATI, log_gp120_ATI)
dat_log_bama = dat_log_sel %>% 
  select(SHIV.CH505TF.V2_week8:SIVmac251.Nef_ATI)
dat_log_sel = cbind(dat_log_prev, dat_log_bama)
```

Then run Lasso+Cox PH model.
```{r}
# X = dat_log_sel[,3:34] %>% as.matrix()

# also do a version without BAMA data
X = dat_log_sel[,3:24] %>% as.matrix()

phmod_lasso = glmnet(X, 
                     Surv(dat_log_sel$rebound_time_days_post_ati, 
                          dat_log_sel$observed),
                     family = "cox")
```

Get a "predictor inclusion" ranking table.
```{r, message=FALSE, warning=FALSE}
set.seed(42)

# get inclusion order
pred_in = NULL
coef_sign = NULL

# hard stop when #(active pred.) >= 7
for(l in phmod_lasso$lambda){
  coefficients = coef(phmod_lasso, s = l)
  active_index = which(coefficients != 0)
  active_coefficients = coefficients[active_index]
  active_predictors = attr(coefficients,"Dimnames")[[1]][active_index]
  
  if(length(active_index)>7){ break }
  
  #cat(active_predictors, "\n")
  
  if(any(!active_predictors %in% pred_in)){
    #cat('Added new predictors! Now active:', active_predictors, '\n')
    
    new_index = which(!active_predictors %in% pred_in)
    pred_in = c(pred_in, active_predictors[new_index])
    new_coef_sign = ifelse(active_coefficients[new_index] > 0, 
                           "+", "-")
    coef_sign = c(coef_sign, new_coef_sign)
  }
}

# focus on those that are eventually selected...

#coef_sign = coef_sign[pred_in %in% active_predictors]
#pred_in = pred_in[pred_in %in% active_predictors]

# rebound effect
rebound_effect = sapply(coef_sign, 
                        function(x) ifelse(x=="+","accelerate","delay")) %>%
  as.vector()

# multivariate concordance (C-statistic)
Response = "Surv(rebound_time_days_post_ati, observed)"

Cum_C_stats = NULL
for(i in 1:length(pred_in)){
  covars = pred_in[1:i]
  f = as.formula(paste(Response,paste(covars,collapse = "+"),sep = " ~ "))
  mod = coxph(f, data=dat_log_sel)
  C_covars = mod$concordance['concordance'] %>% as.numeric()
  Cum_C_stats = c(Cum_C_stats, C_covars)
}

# leave-one-out CV error
cv_phmod_lasso3 = cv.glmnet(X, 
                            Surv(dat_log_sel$rebound_time_days_post_ati, 
                                 dat_log_sel$observed),
                            family = "cox", nfolds = 10, keep = T)
pred_in2 = NULL
LOO_deviance = NULL

# again, hard stop when #(actiev pred.) > 7
for(l in phmod_lasso$lambda){
  coefficients = coef(phmod_lasso, s = l)
  active_index = which(coefficients != 0)
  #active_coefficients = coefficients[active_index]
  active_predictors = attr(coefficients,"Dimnames")[[1]][active_index]
  
  #cat(active_predictors, "\n")
  if(length(active_index)>7){ break }
  
  if(any(!active_predictors %in% pred_in2)){
    new_index = which(!active_predictors %in% pred_in2)
    pred_in2 = c(pred_in2, active_predictors[new_index])
    
    l_index = which(cv_phmod_lasso3$lambda == l)
    LOO_deviance = c(LOO_deviance, cv_phmod_lasso3$cvm[l_index])
  }
}

#LOO_deviance = LOO_deviance[pred_in2 %in% active_predictors]
#pred_in2 = pred_in2[pred_in2 %in% active_predictors]
```

Put together a summary table.
```{r}
predictor_inclusion = data.frame(Inclusion_Rank = c(1:length(pred_in)),
                                 Predictor = pred_in,
                                 #Coefficient = coef_sign,
                                 Rebound_Effect = rebound_effect,
                                 Cum_concordance = Cum_C_stats,
                                 LOOCV_deviance = LOO_deviance)
predictor_inclusion
```

Save the table as txt file for Word use.
```{r}
write.table(format(predictor_inclusion, digits=3), 
            'predictor_inclusion_table_wo_BAMA.txt',
            quote=FALSE, row.names = FALSE,
            sep = ',')
```


KM curves to visualize the effects.



## Test the significance of all BAMA variables
For each BAMA variable, we use (1) value at ATI, and (2) difference between week 8 and ATI. Only adjust for `Sex`, `A01 type`, and `Dosage`.

```{r}
bama.vars = names(bama.dat)
sel_col = str_detect(bama.vars, '(diff)|(ATI)')
dat = cbind(dat_log[,c(1,4,5,43,44,46)], bama.dat[,sel_col])
```

Run a Cox PH model for each variable and get p values.
```{r}
# try not including `Sex` (or `A01`) with the GP140 stuff
# - there seems to be some numeric problems
p_vals = NULL
Response = "Surv(rebound_time_days_post_ati, observed)"
f_st = paste(Response, 'A01 + Sex + Dosage', sep=' ~ ')
#f_st = paste(Response, 'A01 + Dosage', sep=' ~ ')
#f_st = paste(Response, 'Sex + Dosage', sep=' ~ ')
bamas = bama.vars[sel_col]
for(i in seq(1,length(bamas),by=2)){
  sel_var = bamas[c(i,i+1)]
  if(str_detect(sel_var[1],'gp140')){
    f_st1 = paste(Response, 'A01 + Dosage', sep=' ~ ')
    f = as.formula(paste(f_st1, paste(sel_var,collapse = ' + '),sep='+'))
  }else{
    f = as.formula(paste(f_st, paste(sel_var,collapse = ' + '),sep='+'))
  }
  phm = coxph(f, data=dat)
  this.ps = summary(phm)$coefficients[,5][sel_var]
  p_vals = c(p_vals, this.ps)
}
```

Inspect the p-values, and also look at the q-values (FDRs). Here we only show the 10 smallest.
```{r}
#p_vals
sort(p_vals)[1:10]
```

```{r}
q_vals = p.adjust(p_vals, method = 'fdr')
sort(q_vals)
```

The model with `p27`.
```{r}
m_p27 = coxph(Surv(rebound_time_days_post_ati, observed) ~
                A01 + Sex + Dosage + SIVmac239.p27_ART_diff + 
                SIVmac239.p27_ATI,
              data = dat)
summary(m_p27)
```

Put together a summary table of p-values and q-values (only the top ones with smallest p-vals).
```{r}
tops = order(p_vals)[1:6]
bama_tabs = data.frame(Predictor = bamas[tops],
                       p.values = p_vals[tops],
                       q.values = q_vals[tops])
write.table(format(bama_tabs, digits=3), 
            'BAMA_p_values.txt',
            quote=FALSE, row.names = FALSE,
            sep = ',')
```


