---
title: "B2 Rebound Analysis"
author: "Fan Bu"
date: "6/11/2020"
output: slidy_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_knit$set(root.dir = "~/Documents/Research_and_References/HIV_rebound_summer2020/")
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.align = "center")
```

```{r setup 2, echo=FALSE, message=F, warning=F}
library(tidyverse)
library(survival)
library(survminer)

#setwd("~/Documents/Research_and_References/HIV_rebound_summer2020/")
```

## Overview

- Goal of analysis: identify predictors for **viral rebound** (typically happens after treatment interruption)
- Data: B2 group (8 monkeys, late ART at 8 weeks upon infection)
- Results: 

    - Correlates of time-to-rebound: Antibody neutralization capability & Viral load pre-ART
    - Correlates of post-rebound viral load: Viral load pre-ART & Cell-associated RNA and DNA levels
  
- Methods:
  
    - Cox Porportional Hazards model (a survival)
    - Pearson correlation coefficients (& linear regression)


## Time to rebound data

Time to rebound measured in **days**; one of the animals **didn't** rebound ("censorsed").

```{r load data, echo=FALSE}
dat_log = readRDS("reboundB2_logTrans_withVLAUC.rds")
```

```{r rebound times, echo=FALSE}
knitr::kable(dat_log %>% select(ID = animal_id,
                                rebound_time = rebound_time_days_post_ati,
                                observed))
```

## Precitor candidates

These potential predictors are investigated:

- **Viral load pre ART**: peak viral load, viral load at treatment & **area-under-VL-curve**
- **Antibody response**:
    - ELISA: GP41 and GP120, peak level and level at treatment
    - Neutralization: Point IC50 & Pos AUC measured at 0, 2, 4, 8 weeks post ATI
- **Cell-associated RNA and DNA**: 
    - SHIV RNA copies per million cell RNA: measured in blood, lymph node and rectal biopsy at 8 and 56 weeks post infection (RB only at 56 weeks)
    - SHIV DNA copies per million cell DNA: measured in blood, lymph node and rectal biopsy at 8, 16, 36 and 56 weeks post infection (no RB at 8 weeks)


## Survival function and Cox Proportional Hazards (PH) model
Survival function:
$$
S(t) = \text{Pr}(\text{survive past time t)}
$$

Model the Hazard function ($H(t) = 1- S(t)$) as: 
$$
H(t;Z_i) = H_0(t) \exp(\beta^T Z_i),
$$
where $Z_i$ is covariates (predictors) for animal $i$.

Below is the survival curve for this data set.

```{r basic KM curve, echo=FALSE}
simple_KM = survfit(Surv(rebound_time_days_post_ati, observed)~1, data=dat_log)
# plot(simple_KM)
## prettier version
ggsurvplot(simple_KM, data = dat_log, 
           censor.shape="|", censor.size = 4,
           size = 1.5, palette = c("#2E9FDF"),
           conf.int = TRUE,
           ggtheme = theme_bw())
```


## Model selection: single predictor

The "best" univariate Cox PH model is selected via comparing

- "concordance": how the predictor agrees with time-to-rebound
- p-values of significance tests (mainly Likelihood Ratio Test)
- "AIC" & "BIC": model goodness-of-fit criteria

Best predictor: 

- **Area-under-curve (positive part) for antibody neutralization**
- The higher `Pos AUC` is, the lower the hazard is
- Higher antibody neutralization ability **delays** rebound

```{r univariate Cox PH model, echo=F, fig.height=4}
## 1) predictor: pos_auc_0_weeks_post_ATI
phmod_auc0 = coxph(Surv(rebound_time_days_post_ati, observed) ~ pos_auc_0_weeks_post_ATI,
                   data = dat_log)

# summary(phmod_auc0)
# beta = cbind(coef(phmod_auc0), confint(phmod_auc0))
# attr(beta, "dimnames")[[2]][1] = "estimate"
# beta

### 1.a: baseline survival curve (at the MEAN value of pos_auc_0_weeks_post_ATI)
### (should be the same as the plain KM curve...?)
# ggsurvplot(survfit(phmod_auc0, data=dat_log), palette = c("#2E9FDF"),
#            ggtheme = theme_bw())

### 1.b: survival curves at certain representative values of pos_auc_0_weeks_post_ATI
# sort(dat_log$pos_auc_0_weeks_post_ATI)
# [1] 0.1097 0.1474 0.1511 0.2184 0.2512 0.2660 0.2887 0.2959 0.3293 0.4372
# mean(dat_log$pos_auc_0_weeks_post_ATI)
# [1] 0.24949
auc0_values = c(0.15, 0.25, 0.3, 0.4) # 0.25 is approximately the mean
#auc0_values = quantile(dat_log$pos_auc_0_weeks_post_ATI, c(0.25,0.5,0.75))

auc0_fit = survfit(phmod_auc0, 
                   newdata = data.frame(pos_auc_0_weeks_post_ATI = auc0_values))
ggsurvplot(auc0_fit, conf.int = FALSE, 
           data = dat_log,
           legend = "right",
           legend.title = "POS AUC \n0 weeks\npost ATI",
           # legend.labs=c("AUC_0_weeks=0.15", 
           #               "AUC_0_weeks=0.25(mean)",
           #               "AUC_0_weeks=0.30",
           #               "AUC_0_weeks=0.40"),
           legend.labs=c("0.15", 
                         "0.25(mean)",
                         "0.30",
                         "0.40"),
           ggtheme = theme_bw(base_size = 14))
# the confidence bands are HUGE though...

```

## Model selection: two predictors

The "best" bivariate model is selected in similar ways.

Best predictors:

- **Area-under-curve (positive part) for antibody neutralization**
- **Peak viral load** pre ART (averaged between two maximum measurements)
- A higher `Pos AUC` **delays** rebound
- A higher `peak viral load` **accelerates** rebound

## Analysis of post treatment interruption viral load levels

Two measures of "post ATI viral load level" are used (both in log10 scale):

- Peak viral load after rebound, and 
- Ared-under-VL-curve after ATI

```{r post ATI viral load, echo=FALSE, warning=FALSE, message=FALSE}
VL = readRDS("B2_viral_load.rds")

# # visualize by animal IDs
# ggplot(data=VL, aes(x=week_infection, y=log_viral_load)) + 
#   geom_vline(xintercept = c(8,62), size=1) +
#   geom_line(aes(color=animal_id)) +
#   labs(x="Weeks post infection", 
#        y="Log viral load",
#        color="Animal ID")+
#   scale_x_continuous(breaks = c(0,8,25,50,62,75)) +
#   theme_bw(base_size = 14)

# visualize only the part after ATI
VL_post = VL %>% filter(days_post_ati >= 0)

# load rebound time data as well (mark up rebound time in plots)
dat_log = readRDS("reboundB2_logTrans_withVLAUC.rds")
rebound_times = dat_log %>% 
  select(animal_id, rebound_time_days_post_ati)
rebound_times$rebound_time_days_post_ati[rebound_times$animal_id=="RQc19"] = NA

ggplot(data=VL_post, aes(x=days_post_ati, y=log_viral_load)) +
  geom_hline(yintercept = log(60, base=10), 
             size = 0.5, linetype = "dashed") +
  geom_vline(aes(xintercept = rebound_time_days_post_ati),
             data = rebound_times, 
             size = 0.5, linetype = "dashed") +
  geom_line(aes(color=animal_id)) +
  geom_point(aes(color=animal_id), size=1) +
  labs(x="Days post ATI", 
       y="Viral load (log10)",
       color="Animal ID")+
  #scale_x_continuous(breaks = c(0,8,25,50,62,75)) +
  scale_y_continuous(limits = c(0,8)) +
  theme_bw(base_size = 14)+
  facet_wrap(~animal_id, ncol = 5)

```

Pearson correlation coefficients are calculated between each of the viral load measures and each of the potential predictors. Predictors with very high positive correlation include:

- viral load at treatment
- cell-associated DNA & RNA level at treatment
- cell-associated DNA & RNA level during treatment


## Next steps

- Analysis with more potential predictors included
    * Animal MHC class (A01 positive or negative)
    * Cell counts (claimed to be associated with time-to-rebound and post-rebound viral level by some papers)
- Adjust for multiple testing (control false discovery rate)
- Conduct multivariate Cox Regression with sparsity regularization (small $n$, large $p$)

