---
title: 'B2 Rebound Analysis: More'
author: "Fan Bu"
date: "6/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Research_and_References/HIV_rebound_summer2020/")
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.align = "center")
```

```{r load packages, echo=FALSE, message=F, warning=F}
library(tidyverse)
library(survival)
library(survminer)
library(glmnet)
```

```{r load data, echo=FALSE}
dat_log = readRDS("reboundB2_logTrans_CellCounts_A01.rds")
```


## Checking out more predictors for time-to-rebound

### 1. Animal MHC class ("A01" positive/negative)

Below is the Kaplan-Meier curve stratified by "A01" classification. There isn't any visible difference between the two survival curves.

```{r KM curve A01, echo=FALSE}
# 1.1 KM curve stratified by A01 classification
A01_KM = survfit(Surv(rebound_time_days_post_ati, observed)~A01, data=dat_log)
## prettier versions
ggsurvplot(A01_KM, data = dat_log, 
           censor.shape="|", censor.size = 4,
           size = 1.5, palette = c("#E7B800", "#2E9FDF"),
           conf.int = FALSE,
           ggtheme = theme_bw())
```

To further validate, a univariate Cox Proporational Hazards model is fit using "A01" as the predictor, and there isn't a significant effect. 

```{r Cox PH with A01}
phmod_A01 = coxph(Surv(rebound_time_days_post_ati, observed)~A01, 
                   data = dat_log)
summary(phmod_A01)
```

### 2. Cell counts

Absolute CD4 and CD8 cell counts (estimated by flow data + CBC, provided by Veronica) are obtained at Weeks 0, 4, 8. Here I'm using the log-transformed counts as potential predictors. 

As it turns out, none of the cell counts has high concordance with rebound times, compared to other predictors tried out before.


## Regularized Cox PH model (Cox regression)

To conduct "automatic" multiple regression and variable selection, I used Cox PH model combined with elastic net ("coxnet", implemented in package `glmnet`); specifically, for the purpose of variable selection, I used Lasso (L1 regularization, i.e., $\alpha=1$ for elastic net).

I did the following things:

1. Fit "coxnet" models with a series of $\lambda$ (regularization hyper-parameter) values, where the covaraites include:

    - **Viral load pre ART**: peak viral load, viral load at treatment & **area-under-VL-curve**
    - **Antibody response**:
        - ELISA: GP41 and GP120, peak level and level at treatment
        - Neutralization: Point IC50 & POS AUC measured at 0, 2, 4, 8 weeks post ATI
    - **Cell-associated RNA and DNA**: 
        - SHIV RNA copies per million cell RNA: measured in blood, lymph node and rectal biopsy at 8 and 56 weeks post infection (RB only at 56 weeks)
        - SHIV DNA copies per million cell DNA: measured in blood, lymph node and rectal biopsy at 8, 16, 36 and 56 weeks post infection (no RB at 8 weeks)
    - **Cell counts**: Absolute CD4 and CD8 cell counts at 0, 4, 8 weeks.
    - ("A01" not included)
    
2. Select the $\lambda$ that yields the smallest partial deviance in cross validation and obtain the corresponding model.

3. Produce a "predictor inclusion ranking" table of the predictors based on the "coxnet" modeling results. (Note that CV with 10 data points isn't very reliable; the best thing we can get is probably some "variable importance" ranking.)


Below are the CV plot (partial likelihood deviance vs. $\log(\lambda)$) and the selected predictors and coefficients (different runs can produce different outcomes, so it's unstable).

```{r Coxnet, echo=FALSE}
# 3.1 get the design matrix 
# (ignore A01 for now - can only deal with numeric values)
X = dat_log[,6:42] %>% as.matrix()

# 3.2 fit Lasso (alpha=1)
phmod_lasso = glmnet(X, 
                     Surv(dat_log$rebound_time_days_post_ati, dat_log$observed),
                     family = "cox")

# (cross validation)
## (using partial likelihood, 5-fold)
cv_phmod_lasso1 = cv.glmnet(X, 
                           Surv(dat_log$rebound_time_days_post_ati, dat_log$observed),
                           family = "cox", nfolds = 5)
plot(cv_phmod_lasso1)

coefficients <- coef(phmod_lasso, s = cv_phmod_lasso1$lambda.min)
active_index <- which(coefficients != 0)
active_coefficients <-coefficients[active_index]
active_predictors = attr(coefficients,"Dimnames")[[1]][active_index]

### put together a table for this result
cv_res = data.frame(Predictor = active_predictors, 
                    Coefficient = active_coefficients)

knitr::kable(cv_res, digits = 4)
```

Below is the obtained "predictor inclusion ranking" table for the top 9 predictors. 

```{r variable inclusion ranking, echo=F}
# 3.3 Use Lasso results to obtain a "predictor inclusion ranking"
pred_in = NULL
coef_sign = NULL

for(l in cv_phmod_lasso1$lambda){
  coefficients = coef(phmod_lasso, s = l)
  active_index = which(coefficients != 0)
  active_coefficients = coefficients[active_index]
  active_predictors = attr(coefficients,"Dimnames")[[1]][active_index]
  
  if(any(!active_predictors %in% pred_in)){
    new_index = which(!active_predictors %in% pred_in)
    pred_in = c(pred_in, active_predictors[new_index])
    new_coef_sign = ifelse(active_coefficients[new_index] > 0, 
                           "+", "-")
    coef_sign = c(coef_sign, new_coef_sign)
  }
}

## get a vector of "effect on rebound"
## (+: accelerate; -: delay)
rebound_effect = sapply(coef_sign, 
                        function(x) ifelse(x=="+","accelerate","delay")) %>%
  as.vector()

## put together a summary table of this thing
predictor_inclusion = data.frame(Inclusion_Rank = c(1:9),
                                 Predictor = pred_in,
                                 Coefficient = coef_sign,
                                 Rebound_Effect = rebound_effect)
knitr::kable(predictor_inclusion, digits = 4)
```


## Analysis of post-ATI viral load
I also re-explored the post-ATI viral loads, with animal `RQc19` (the one without rebound) included, as well as cell counts data. 

The results did differ from last time. 

```{r load postATI data, echo=F}
dat_log_rebound = readRDS("rebound_B2_with_postATI_VL.rds")
```


Variables that correlate with `log peak VL` include:

- `log_DNA_copies_RB_36`
- `pos_auc_0_weeks_post_ATI`
- `log_point_ic50_0_weekspost_ATI`

```{r post-ATI peak VL correlations, echo=F}
## post peak vs RB DNA copies
ggscatter(data = dat_log_rebound, 
          y = "log_peakVL_postATI", x = "log_DNA_copies_RB_36",
          add = "reg.line",
          add.params = list(color = "blue", fill = "lightgray"),
          conf.int = T,
          cor.coef = T)

## post peak vs "pos_auc_0_weeks_post_ATI"
ggscatter(data = dat_log_rebound, 
          y = "log_peakVL_postATI", x = "pos_auc_0_weeks_post_ATI",
          add = "reg.line",
          add.params = list(color = "blue", fill = "lightgray"),
          conf.int = T,
          cor.coef = T)

## post peak vs "pos_auc_0_weeks_post_ATI"
ggscatter(data = dat_log_rebound, 
          y = "log_peakVL_postATI", x = "log_point_ic50_0_weekspost_ATI",
          add = "reg.line",
          add.params = list(color = "blue", fill = "lightgray"),
          conf.int = T,
          cor.coef = T)
```

Variables that correlate with `VL AUC` include:

- `log_DNA_copies_Blood_36`      
- `log_DNA_copies_RB_36`        
- `log_RNA_copies_LN_8`
- etc. (followed by some other DNA & RNA measurements, and VL at treatment)

```{r post-ATI VL AUC correlations, echo= FALSE}
## post VL AUC vs log_DNA_copies_Blood_36
ggscatter(data = dat_log_rebound, 
          y = "log_VL_AUC_postATI", x = "log_DNA_copies_Blood_36",
          add = "reg.line",
          add.params = list(color = "blue", fill = "lightgray"),
          conf.int = T,
          cor.coef = T)

## post VL AUC vs log_DNA_copies_RB_36
ggscatter(data = dat_log_rebound, 
          y = "log_VL_AUC_postATI", x = "log_DNA_copies_RB_36",
          add = "reg.line",
          add.params = list(color = "blue", fill = "lightgray"),
          conf.int = T,
          cor.coef = T)
```


I also compared subgroups of positive `A01` vs. negative `A01` (5 monkeys each). 

For "Area-under-VL-curve", it looks like some pattern is there...
```{r VL AUC boxplot, echo=F}
## log VL AUC
ggplot(data=dat_log_rebound, 
       aes(x=A01, y=log_VL_AUC_postATI)) +
  geom_boxplot() + 
  theme_bw()
```

But is there??
```{r VL AUC scatterplot, echo=F}
## (as points...)
ggplot(data=dat_log_rebound, 
       aes(x=A01, y=log_VL_AUC_postATI)) +
  geom_point() + 
  theme_bw()
```

A Wilcoxon rank sum test (equivalent to the Mann-Whitney test in this case) is performed (despite the tiny sample size). There is no strong evidence to reject the null hypothesis (p-value = 0.155).
