---
title: "B2 rebound survival analysis"
author: "Fan Bu"
date: "6/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/Research_and_References/HIV_rebound_summer2020/")
knitr::opts_chunk$set(fig.width=8, fig.height=6, fig.align = "center")
```

```{r setup 2, echo=FALSE, message=F, warning=F}
library(tidyverse)
library(survival)
library(survminer)

#setwd("~/Documents/Research_and_References/HIV_rebound_summer2020/")
```

## Overview

The goal of the analysis is to study influential predictors for time to rebound. The data analyzed here are obtained in the B2 group experiments. 

These potential predictors are investigated:

- **Viral load pre ART**: peak viral load & viral load at treatment
- **Antibody response**:
    - ELISA: GP41 and GP120, peak level and level at treatment
    - Neutralization: Point IC50 & POS AUC measured at 0, 2, 4, 8 weeks post ATI
- **Cell-associated RNA and DNA**: 
    - SHIV RNA copies per million cell RNA: measured in blood, lymph node and rectal biopsy at 8 and 56 weeks post infection (RB only at 56 weeks)
    - SHIV DNA copies per million cell DNA: measured in blood, lymph node and rectal biopsy at 8, 16, 36 and 56 weeks post infection (no RB at 8 weeks)


```{r load data, echo=FALSE}
dat_log = readRDS("reboundB2_logTrans_withDNA.rds")
```

Among the 10 monkeys in group B2, 9 experienced rebound (except for RQc19, labelled as "0" in the "observed" column). Below is a summary table of their rebound times (measured in days).

```{r rebound times, echo=FALSE}
knitr::kable(dat_log %>% select(ID = animal_id,
                                rebound_time = rebound_time_days_post_ati,
                                observed))
```

Below is the basic Kaplan Meier curve of the rebound times. This curve shows the probability of surviving past any time point; that is, a point (t, p) on the curve means that a monkey has a $p \times 100\%$ chance of **not** having rebound till $t$ days. 

```{r basic KM curve, echo=FALSE}
simple_KM = survfit(Surv(rebound_time_days_post_ati, observed)~1, data=dat_log)
# plot(simple_KM)
## prettier version
ggsurvplot(simple_KM, data = dat_log, 
           censor.shape="|", censor.size = 4,
           size = 1.5, palette = c("#2E9FDF"),
           conf.int = TRUE,
           ggtheme = theme_bw())
```


## Predictors transformations and correlations

### Predictor distributions and potential transformations

We can notice that

- all of the potential predictors are positive values, and 
- most of the variables have large ranges and dispersed values

Therefore, the predictors are log-transformed. 

(Moreover, we can check the "martingale residuals" to select proper functional forms for the predictors so that linearity would be satisfied for the Cox Porportional Hazards model. Some level of non-linearity is seen, and log-transformation does make the residuals seem more linear, but the small sample size doesn't allow much certainty on this.)

Further, another representation of the "peak viral load" is created: `peak_vl_2` = average of **top 2** viral load measurements pre ART. This is used in the hope to compensate for possible measurement noise in the "peak" viral load.


### Correlations between predictors

Pairwise scatterplots are created to study significant correlations between potential predictors. (Basically, predictors that are hightly correlated shouldn't be used together in a model.)

Below are the main findings:

  - GP41 concentration at peak and at treatment almost perfectly correlated
  - peak viral load highly correlated with GP120 at treatment and at peak
  - GP120 concentration at peak and at treatment also highly correlated
  - the point IC50 measurements and POS AUC values are all highly correlated, especially those measurements acquired consecutively as well as point IC50 and AUC in the same week
  - viral load at treatment highly correlated with "RNA blood 8 weeks", "RNA LN 8 weeks" and "DNA blood 8 weeks"
  - peak VL highly correlated with "RNA blood 8 weeks"
  - "RNA blood 8 weeks" and "RNA LN 8 weeks" highly correlated
  - "DNA blood 8 weeks" and "DNA blood 16 weeks" & "DNA LN 8 weeks" and "DNA LN 36 weeks" highly correlated


<!-- (also include the RNA-DNA plot; ask why the week 56 counts do not correlate) -->


## Univariate survival analysis 

### Model description: Cox Porportional Hazards model
Suppose subject $i$ has covariates $Z_i$, then the hazard function ($=1-S(t)=1-Pr(\text{survive past time } t)$) for $i$ is

$$
H(t;Z_i) = H_0(t) \exp(\beta^T Z_i),
$$
where $H_0(t)$ is the baseline hazard function. Throughout this analysis, we use centralized (i.e, de-meaned) covariates $Z_i$. 

### Non-parametric check: concordance ("C-statistic")

Simply put, concordance is a metric of agreement between a survival model (or a predictor) and a binary survival outcome; its interpretation is very similar to that of AUC (area under the ROC curve) for binary classification. 

Suppose a model assigns score function $g(Z_i)$ for a subject $i$ with covariates $Z_i$, then the concordance (also referred to as the "C-Statistic") is defined as
$$
Pr(g(Z_i) > g(Z_j) \mid T_i > T_j),
$$
for any two subjects $i$ and $j$ with event time $T_i$ and $T_j$ respectively. 

Of course, if we set $g$ as the identity function, then we can measure the concordance w.r.t. to the predictors themselves. Below we show the top 5 predictors with highest concordance in the dataset.

```{r concordance, echo=F}
Response = "Surv(rebound_time_days_post_ati, observed)"
All_covars = names(dat_log)[6:35]

# 4.1 check concordance, the c-statistic

C_stats = NULL

for(v in All_covars){
  f = as.formula(paste(Response,v,sep = " ~ "))
  C_v = concordance(f, data=dat_log, timewt = "n")$concordance
  C_stats = c(C_stats, C_v)
}

C_stats = data.frame(Predictor = All_covars, Concordance = C_stats)

## show it with descending rank
knitr::kable(
  C_stats %>% arrange(desc(Concordance)) %>% head(.,n=5),
  digits = 5
)

```


### Model selection

To select the best univariate model, an exhaustive search is carried out by fitting a Cox PH model using each of all the potential predictors. 

Then select the most promising predictor(s) by comparing AIC, as well as the p-values obtained by 

- log-rank test
- Wald test
- likelihood ratio test (said to have best behavior for small sample)

At the end, the univariate model with `pos_auc_0_weeks_post_ATI` has the smallest AIC and the smallest p-value across the tests ($p=0.0024$ for LRT).

### The chosen model

`pos_auc_0_weeks_post_ATI` has a significant negative effect on the hazard function; that is, **a higher AUC at ATI delays rebound**. 

```{r univariate Cox PH model, echo=F, fig.height=4}
## 1) predictor: pos_auc_0_weeks_post_ATI
phmod_auc0 = coxph(Surv(rebound_time_days_post_ati, observed) ~ pos_auc_0_weeks_post_ATI,
                   data = dat_log)

# summary(phmod_auc0)
beta = cbind(coef(phmod_auc0), confint(phmod_auc0))
attr(beta, "dimnames")[[2]][1] = "estimate"
beta

### 1.a: baseline survival curve (at the MEAN value of pos_auc_0_weeks_post_ATI)
### (should be the same as the plain KM curve...?)
# ggsurvplot(survfit(phmod_auc0, data=dat_log), palette = c("#2E9FDF"),
#            ggtheme = theme_bw())

### 1.b: survival curves at certain representative values of pos_auc_0_weeks_post_ATI
# sort(dat_log$pos_auc_0_weeks_post_ATI)
# [1] 0.1097 0.1474 0.1511 0.2184 0.2512 0.2660 0.2887 0.2959 0.3293 0.4372
# mean(dat_log$pos_auc_0_weeks_post_ATI)
# [1] 0.24949
auc0_values = c(0.15, 0.25, 0.3, 0.4) # 0.25 is approximately the mean
#auc0_values = quantile(dat_log$pos_auc_0_weeks_post_ATI, c(0.25,0.5,0.75))

auc0_fit = survfit(phmod_auc0, 
                   newdata = data.frame(pos_auc_0_weeks_post_ATI = auc0_values))
ggsurvplot(auc0_fit, conf.int = FALSE, 
           data = dat_log,
           legend = "right",
           legend.title = "POS AUC \n0 weeks\npost ATI",
           # legend.labs=c("AUC_0_weeks=0.15", 
           #               "AUC_0_weeks=0.25(mean)",
           #               "AUC_0_weeks=0.30",
           #               "AUC_0_weeks=0.40"),
           legend.labs=c("0.15", 
                         "0.25(mean)",
                         "0.30",
                         "0.40"),
           ggtheme = theme_bw(base_size = 14))
# the confidence bands are HUGE though...

```


### Model validation
The following steps are conducted to validate the model/check model assumptions:

1. Testing on Schoenfeld residuals to validate "proportional hazards" assumption
    - There isn't a clear pattern of residuals accross time, and p-value = 0.6772, which both suggest that there is no strong evidence against the PH assumption.
2. Examining deviance residuals to identify influential observations/outliers
    - Positive residual = subject died too soon
    - Negative residual = subject lived too long
    - There isn't any residual too far from zero, suggesting that all the subjects follow similar patterns


## Bivariate survival analysis

### Model selection method

One natural way of extending the previously chosen univariate model is to find the best "addtional" predictor. Thus, an exhaustive search on the second predictor is carried out; here, we not only include linear terms but also try out a model with **an additional interaction term** of the two predictors. 

As mentioned before, `pos_auc_0_weeks_post_ATI` is highly positively correlated with all the other antiboby neutralization measurements, so all the other "POS AUC" and "point IC50" variables are not included in the candidate set. 

The best model is selected as the model with the smallest AIC.

(An exhaustive search among all possible bivariate models is also done as a safety measure; the resulting best model is the same.)

### The chosen bivariate model

The chosen model consists of two predictors:

- `pos_auc_0_weeks_post_ATI`
- `log_peak_vl_2` (the newly created peak viral load measure, in log-scale)
- (there is no interaction term)

To summarize:
- "POS AUC" upon ATI **delays** rebound (negative effect on hazard)
- peak viral load **accelerates** rebound (positive effect on hazard)

```{r bivariate model, echo=F, fig.height=4}
f_auc0 = "Surv(rebound_time_days_post_ati, observed) ~ pos_auc_0_weeks_post_ATI"


## look at this new bivariate model
phmod_auc0_peakVL2 = coxph(update(as.formula(f_auc0), ~ . + log_peak_vl_2), 
                          data = dat_log)
# summary(phmod_auc0_peakVL2)

### Summary:
### Concordance: 0.932 (same as before with log_peak_vl)
### pos_auc_0_weeks_post_ATI: negative effect on hazard (delays rebound)
### log_peak_vl: positive effect on hazard (accelerates rebound)
### ALTHOUGH none of the effects is significantly non-zero (same as before)

# confint(phmod_auc0_peakVL2)
#                               2.5 %   97.5 %
# pos_auc_0_weeks_post_ATI -166.849287 12.46821
# log_peak_vl_2              -1.521619 13.41001

## add: a prettier summary of estimates and CIs
beta2 = cbind(coef(phmod_auc0_peakVL2), 
              confint(phmod_auc0_peakVL2))
attr(beta2, "dimnames")[[2]][1] = "estimate"
beta2


## a) baseline survival curve (at the mean values)
# ggsurvplot(survfit(phmod_auc0_peakVL2, data=dat_log), 
#            palette = c("#2E9FDF"),
#            ggtheme = theme_bw())
# HUGE confidence intervals

## b) survival curves at representative values of each variable
## i) fix peak VL at mean, vary AUC_0_week
# mean(dat_log$log_peak_vl_2)
# [1] 5.8631
auc0_values = c(0.15, 0.25, 0.3, 0.4) # 0.25 is approximately the mean

auc0_fit = survfit(phmod_auc0_peakVL2, 
                   newdata = data.frame(pos_auc_0_weeks_post_ATI = auc0_values,
                                        log_peak_vl_2 = mean(dat_log$log_peak_vl_2)))
ggsurvplot(auc0_fit, conf.int = FALSE, 
           data = dat_log,
           legend = "right",
           legend.title = "POS AUC \n0 weeks\npost ATI",
           legend.labs=c("0.15", 
                         "0.25(mean)",
                         "0.30",
                         "0.40"),
           caption = "Fix log_peak_VL_2 at 5.86 (mean)",
           ggtheme = theme_bw(base_size = 14))

## ii) fix AUC_0_week at mean, vary log peak VL 2
# sort(dat_log$log_peak_vl_2)
# [1] 4.366423 5.196136 5.597164 5.651816 5.812189 6.008954 6.103290
# [8] 6.469425 6.483587 6.942020
# mean(dat_log$pos_auc_0_weeks_post_ATI)
# [1] 0.24949
peakVL_values = c(4.5,5.2,5.8,6.5) # 5.8 is approximately the mean value

peakVL_fit = survfit(phmod_auc0_peakVL2, 
                     newdata = data.frame(pos_auc_0_weeks_post_ATI = mean(dat_log$pos_auc_0_weeks_post_ATI),
                                          log_peak_vl_2 = peakVL_values))
ggsurvplot(peakVL_fit, conf.int = FALSE, 
           data = dat_log,
           legend = "right",
           legend.title = "Peak viral load 2\n(log-scale)",
           legend.labs=c("4.5", 
                         "5.2",
                         "5.8(mean)",
                         "6.5"),
           caption = "Fix pos_auc_0_weeks_post_ATI at 0.25 (mean)",
           ggtheme = theme_bw(base_size = 14))
```

### Model validation
Same validation procedures as described above are employed to check the bivariate model. Again, there is no strong evidence against the PH assumption, and there doesn't seem to be any outlier. 

## Discussions

Below are some of my questions:
1. Do we expect the CA-RNA and CA-DNA counts to be highly correlated? We can only see noticeable positive correlation on week 8 (post infection) measurements, but not at all for the following weeks, especially not on week 56. 

2. Might it be promising to incorporate the **full viral load trajectory** (pre ART) into the model? Using all the measurements as predictors would be unrealistic (small sample size!), so it may be possible to learn some "summary statistics" of the viral load trajectory?
    a. fit each individual curve with a simple function and use the learned parameters as predictors
    b. use a latent factor model to learn lower-dimensional representations of the viral load time series, and then use the learned features as predictors
    c. etc.?

3. Data analysis for other experiment groups?    


